---
title: "Homework 2"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
#suppressPackageStartupMessages(library(KKNN))
#suppressPackageStartupMessages(library(MASS))

```

__Submission Instructions__
Submit the R Markdown (.Rmd) file with code and answers on Brightspace.

## **Question 1: Large Language Model (LLM) for Text Analysis**

Kaggle website has an archive of Donal Trump's Tweets during his first presidency, until 8-Jan-2021. The data contains information on number of retweets, deletion of tweets, device through which tweeted, flagged tweets, favorite tweets, etc. For this question, you will work with this data set. 

```{r}

DTtweets=read.csv("https://raw.githubusercontent.com/babakrezaee/MethodsCourses/refs/heads/master/DataSets/trump_tweets.csv")

# Let's first see the dimension of the dataset
dim(DTtweets)

```
You will use ChatGPT API to analyze the sentiment of Donald Trump’s tweets related to NATO and the European Union (EU).

```{r}
library(dplyr)

# Convert text to lowercase for case-insensitive filtering
DTtweets <- DTtweets %>%
  mutate(text = tolower(text))

# Filter tweets mentioning NATO, EU, and China
nato_tweets <- filter(DTtweets, grepl("nato", text))
eu_tweets <- filter(DTtweets, grepl("eu|european union|europeanunion",text))

# Display counts
nrow(nato_tweets)
nrow(eu_tweets)

```
Your main tasks are:

1. Use ChatGPT API to analyze the sentiment of each tweet, as a categorical (Positive, Neutral, and Negtaive) and continuous variable (from -1 to +1).

2. Compare and interpret the sentiment trends across topics over time. 

3. Analyze whether the popularity of these tweets are associated with their sentiment. Use a regression analysis. tip: first think about your unit of analysis.


## Question 2: Random Forest Regression Using _caret_ package

In class, you learned to use _randomForest_ package to develop a random forest algorithm. In this assignment, you are asked to develop a 5-fold cross-validated random forest model using _caret_ package, which you learned about in [Naive Bayes](https://babakrezaee.github.io/Computational_IR/NaiveBayes.html) session.

Load the dataset:

```{r}

library(readr)

# Load dataset
CHdata <- read_csv("https://raw.githubusercontent.com/babakrezaee/MethodsCourses/master/DataSets/calhouse.csv")

# Check structure of dataset
head(CHdata)
colnames(CHdata)
dim(CHdata)
```

Now, partition the data into train (80%) and test (20%).

```{r}
library(caret)

# Set seed for reproducibility
set.seed(7)

# Shuffle the data
CHdata <- CHdata[sample(nrow(CHdata)), ]

# Create a partition (80% train, 20% test)
trainIndex <- createDataPartition(CHdata$logMedVal, p = 0.8, list = FALSE)

# Split the dataset
CAtrain <- CHdata[trainIndex, ]
CAtest <- CHdata[-trainIndex, ]

# Check dimensions
dim(CAtrain)
dim(CAtest)

```

For _caret_ package, you first need to declare the traincontrol() function, where you specify cross-validation and the number of folds. 

```{r}

# Train Random Forest model using `caret`
control <- trainControl(method = "cv", number = 5, verboseIter = TRUE)  # 5-fold cross-validation

RF_model <- train(
  logMedVal ~ ., 
  data = CAtrain, 
  method = "rf", 
  trControl = control, 
  tuneGrid = expand.grid(mtry = 3),  # Using mtry=3
  ntree = 100  # Number of trees
)

# Print model summary
print(RF_model)

```
Now that you trained your model, we can make Predictions and Evaluate Performance.

```{r}
# Predict on the test set
RFpred <- predict(RF_model, newdata = CAtest)

# Compute RMSE
RMSE <- sqrt(mean((CAtest$logMedVal - RFpred)^2))

cat("RMSE on the test data for Random Forest using caret is:", RMSE, "\n")

```

Here are your tasks for this question:

1. What is the purpose of _verboseIter = TRUE_ in the above code?
2. What does mtry=3 and ntree=100 mean in the Random Forest model?
3. How does increasing ntree (number of trees) impact model performance?
4. Explain how we can find the best mtry here? Discuss your tuning strategy. tip: you can set expand.grid(mtry = (3:5))! IMPORTANT: This is a computationally intensive process!


